Analysing Constraint Grammar with a SAT solver

### 2015-10-14 paper/lrec/problem_desc.tex
more text about the CG encoded in logic stuff

Constraint Grammar (CG, \cite{karlsson1995constraint})
% \citep[CG][]{karlsson1995constraint} 
is a method for disambiguating morphologically analysed text% and adding syntactic or dependency labels
. 
A grammar consists of rules that target specific analyses for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1C det) ;}
\end{itemize}
removes a verb reading from a word which is preceded by an unambiguously (C) tagged determiner.
Given a text, such as

\begin{verbatim}
"<the>"
        "the" det def
"<bear>"
        "bear" noun sg
        "bear" verb pl
\end{verbatim}

the rule will apply to the word \emph{bear}, and remove the analysis \texttt{verb pl}.
Rules are not allowed to remove the last remaining analysis: if  \texttt{verb pl} were the only analysis for \emph{bear}, then the rule would not apply, even if the condition is met.

@@

As CGs grow larger, it gets harder for the grammar writers and users to keep track of all the rules and their interaction.
\cite{voutilainen2004} gives a detailed account about best practices of grammar writing, and efficient use of hand-annotated corpora to aid the grammar development.
\cite{bick2013tuning} presents a method to optimise hand-written CGs using machine learning techniques.
Our toolset is designed to complement the existing approaches in CG analysis and optimisation.
While Bick's experiments show improvements in results, the grammar writer is none the wiser why is the grammar better.
Authors who have a large hand-annotated corpus can also benefit from the technique.
With our method, it is possible to ask, for example, the following questions:

@@

\begin{itemize}
\item Corpus-based methods in manual grammar development \cite{voutilainen2004}, \cite{tyers_reynolds2015}
\item Optimising hand-written grammars \cite{bick2013tuning}
\item Encoding CG in logic \cite{lager98}, \cite{lager_nivre01}, \cite{listenmaa_claessen2015}
\end{itemize}

@@

The connection to the reality has also its limitations. 
% What if instead of firing 500 times, it could have fired 1000 times if it had been placed earlier in the rule sequence.
Or the rule that was wrong in 60 \% of the time,

@@

\paragraph{Things you can get from SAT solver}

>>>>

Constraint Grammar (CG, \cite{karlsson1995constraint})
% \citep[CG][]{karlsson1995constraint} 
is a method for disambiguating morphologically analysed text% and adding syntactic or dependency labels
. 
A grammar consists of rules that target specific analyses for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1C det) ;}
\end{itemize}
removes a verb reading from a word which is preceded by an unambiguously (C) tagged determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<bear>"
        "bear" noun sg
        "bear" verb pl
\end{verbatim}
\end{itemize}
the rule will apply to the word \emph{bear}, and remove the analysis \texttt{verb pl}.
However, if a word has only one remaining analysis, then the rule will not apply, even if the condition is met.
%Rules are not allowed to remove the last remaining analysis: if  \texttt{verb pl} were the only analysis for \emph{bear}, then the rule would not apply, even if the condition is met.

@@

As CGs grow larger, it gets harder for the grammar writers and users to keep track of all the rules and their interaction.
\cite{voutilainen2004} gives a detailed account about best practices of grammar writing and efficient use of hand-annotated corpora to aid the grammar development.
\cite{bick2013tuning} presents a method to optimise hand-written CGs using machine learning techniques.
Our toolset is designed to complement the existing approaches in CG analysis and optimisation.
While Bick's experiments show improvements in results, the grammar writer is none the wiser why is the grammar better.
Authors who have a large hand-annotated corpus can also benefit from the technique.
With our method, it is possible to ask, for example, the following questions:

@@

\begin{itemize}
\item Corpus-based methods in manual grammar development \cite{voutilainen2004}
\item Optimising hand-written grammars \cite{bick2013tuning}
\item Encoding CG in logic \cite{lager98}, \cite{lager_nivre01}, \cite{listenmaa_claessen2015}
\end{itemize}

@@

%The connection to the reality has also its limitations. 
% What if instead of firing 500 times, it could have fired 1000 times if it had been placed earlier in the rule sequence.
%Or the rule that was wrong in 60 \% of the time,

@@

\paragraph{CG encoded in logic}

\cite{lager98} presents a CG-like shallow parsing system encoded in logic, and \cite{lager_nivre2001} continues with a reconstruction of four different formalisms. \cite{listenmaa_claessen2015} presents a working CG implementation.

%The earlier works on logical reconstruction don't envision grammar analysis as one of the use cases,
Grammar analysis is a natural use case, due to some key features of the logical reconstruction.

The straight-forward implementation of a CG compiler goes through the text, applies the rules, and discards analyses at each round, losing all information from the previous round when moving on to the next.
A logic-based implementation models the rules as implications, where the consequent from rule $i$ becomes the antecedent of the rule $i+1$. 
Composing all the clauses gives the final analysis of each word.
What we added to this: at any time, we can stop and ask 
``after applying these rules, can we build a sentence with properties \_\_\_?''

% Application of rules is presented as a composition of clauses

% They model CG as clauses o
% The key elements in the ontology of CG are positions, words and sets of tags.
% Rule order is denoted by predicate $pos^i$(word, [tag]), 
% which denotes the part of speech of a given word after applying the $i^{th}$ rule.
% The rules are modelled as implications, of the form below:
